``Interestingness'' in Mathematical Theory Exploration

### 2019-06-18 PhDThesis/clustering.tex
Lots of clustering changes

To improve the performance of theory exploration tools, we need to avoid
brute-force search in favour of a more informed strategy. This, in turn,
requires more information about the terms in the given signature. One
generally available source of information is a term's definition.

Unfortunately these definitions are over-complicated due to the relatively large
syntax of Haskell, which includes many convenience features (``syntactic
sugar'') and alternative ways to write essentially the same thing (e.g.
\hs{let} vs \hs{where}, \hs{if} vs \hs{case}, \hs{f x = ...} vs
\hs{f = \x -> ...}, etc.). This obscures the simplicity of the lambda calculus
underlying the language, and more importantly it makes syntactic analysis less
useful for our purposes by making distinctions between semantically identical
terms (which will behave identically under theory exploration). For this reason
we perform our analyses on the \emph{GHC Core} language instead.

@@

GHC Core is an intermediate language of the \textsc{GHC} compiler based on
\fc{}, and described in detail in~\cite[Appendix C]{sulzmann2007system}. Core
contains explicit type annotations and coercions, which we omit as they have no
effect on runtime behaviour (they have no ``computational content''). The
resulting sub-set of Core\footnote{As of GHC version 7.10.2.} is shown in
Figure~\ref{fig:coresyntax}.

@@

GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
shows the Core representation of some simple functions. Although the Core is
more verbose, we can see that similar structure in the Haskell definitions gives
rise to similar structure in the Core; for example, the definitions of \hs{odd}
and \hs{even} are identical in both languages, except for the particular
identifiers used. It is this close correspondence which allows us to analyse
Core expressions in place of their more complicated Haskell source.

For simplicity we also avoid using constructors directly, choosing instead to
wrap them with normal functions (e.g. \hs{s x = S x}, known as
$\eta$-expansion), since this eliminates another distinction (constructors vs
functions) and also reduces cross-language differences (e.g. Isabelle treats
constructors in a different way to Haskell).

@@

Raw syntax trees, even in Core form, are unsuitable for many forms of analysis
due to their recursive structure, use of discrete labels and heavy reliance on
references. These are unsuitable for many machine learning algorithms, which
instead require a fixed-size vector of continuous numbers known as a
\emph{feature vector}.

The process of turning raw data into feature vectors is called \emph{feature
  extraction}, and we adapt the methodology of \emph{recurrent clustering}
proposed in~\cite{DBLP:journals/corr/HerasK14,heras2013proof}, since this
handles recursive tree structures, resolves references and produces numeric
vectors. We suggest a new recurrent clustering and feature extraction algorithm
for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
and differences to the existing approaches of ML4PG (Coq) and ACL2(ml) (ACL2).

@@

This is a standard setup, supported by off-the-shelf tools; in particular we use
the \hPackage{k-means} Haskell package.

@@

We chose k-means because it is a standard approach and we are more concerned
with the application of feature extraction to Haskell and its use in theory
exploration, rather than precise tuning of learning algorithms. There are many
other clustering algorithms we could use, such as \emph{expectation
  maximisation}~\footnote{In fact, k-means is very similar to
  expectation-maximisation, as it alternates between an \emph{expectation step}
  (finding the mean value of each cluster) and a \emph{maximisation step}
  (assigning points to the cluster they're most similar to; or alternatively,
  \emph{minimising} the distance of each point to the centre of its cluster, as
  per equation \ref{eq:kmeansobjective}).}, but experiments with ML4PG have
shown little difference in their results; in effect, the quality of the features
is the bottleneck to learning, so there is no reason to avoid a fast algorithm
like k-means.

In any case there are many conservative improvements to the standard k-means
algorithm, which could be applied to our setup. For example, a more efficient
approach like \emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make
larger input sizes more practical to cluster, especially since recurrent
clustering invokes k-means many times. The \emph{k-means++}
approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
select the ``seed'' values for the first timestep, and the \emph{x-means}
algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use.

\subsection{Recurrent Clustering}

\emph{Recurrent clustering} has been implemented in the
ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
ACL2(ml) for ACL2~\cite{heras2013proof}. Whilst both transform syntax trees into
matrices, the algorithm of ML4PG most closely resembles ours as it assigns tokens directly to matrix elements. In contrast,
the matrices produced by ACL2(ml) \emph{summarise} information about the tree;
for example, one column counts the number of variables appearing at each tree
level, others count the number of function symbols which are nullary, unary,
binary, etc. Whilst it may be interesting to contrast our current algorithm with
an alternative based on that of ACL2(ml), it is unclear how such summaries could
be extended to include types, which seems the next logical step for our
approach. The ML4PG algorithm extends trivially, by using (term, type) pairs
instead of just terms.

ACL2(ml) reasons by analogy: finding theorem statements which are similar to the
current goal, and attempting to prove the goal in a similar way. In particular,
the lemmas used to prove a theorem are mutated by substituting symbols for those
which appear in the same cluster. For example, if \texttt{plus} and
\texttt{multiply} are clustered together, and we are trying to prove a goal
involving \texttt{multiply}, then ACL2(ml) might consider an existing theorem
involving \texttt{plus}. The lemmas used to prove that theorem will be mutated,
for example replacing occurrences of \texttt{plus} with \texttt{mult}, in an
attempt to prove the goal.

\iffalse
% TODO: Move this to the end of the chapter?
Whilst we do not currently reason by analogy, this is an interesting area for
future work in theory exploration: given a set of theorems relating particular
terms, we might form conjectures regarding similar terms found through
clustering.
\fi

\subsection{Rose Trees}\label{sec:rosetree}

The $toTree$ function shown below transforms Core expressions, described in
\S~\ref{sec:core}, to rose trees. We follow the presentation
in~\cite{blundell2012bayesian} and define rose trees recursively as follows: $T$
is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and
$T_i$ are rose trees.

The recursive definition is mostly routine; each repeated element (shown as
$\dots$) has an example to indicate their handling, e.g. for $\CRec$ we apply
$toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since constructors
have no internal structure for us to compare. We also ignore values from
$\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of $\phi$,
and we conjecture that the effect of such ``magic values'' on clustering real
code is low.

\begin{figure}
  \begin{align*}\label{fig:totree}
    toTree(e) &=
    \begin{cases}
      (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
      (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
      (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
      (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
      (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
      (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
      (\feature{\CType})                                                & \text{if $e = \CType$} \\
      (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
      (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
      (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
      (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
      (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
      (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
      (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
      (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
      (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
      (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
      (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
      (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
    \end{cases}
  \end{align*}
\end{figure}

\subsubsection{Expressions to Vectors}
\label{sec:expressionstovectors}

Our recurrent clustering algorithm is based on k-means clustering, which
considers the elements of a feature vector to be \emph{orthogonal}. Hence we
must ensure that similar expressions not only give rise to similar numerical
values, but crucially that these values appear \emph{at the same position} in
the feature vectors. Since different patterns of nesting can alter the ``shape''
of expressions, simple traversals (breadth-first, depth-first, post-order, etc.)
may cause features from equivalent sub-expressions to be mis-aligned. For
example, consider the following expressions, which represent pattern-match
clauses with different patterns but the same body (\hs{\vlocal{y}}):

@@

If we traverse these expressions in breadth-first order, converting each token
to a feature using $\phi$ and padding to the same length with $0$, we would get
the following feature vectors:

@@

The method we use to ``flatten'' expressions, described below, is a variation of
breadth-first traversal which pads each level of nesting to a fixed size $c$
(for \emph{columns}). This doesn't guarantee alignment, but it does prevent
mis-alignment from accumulating across different levels of nesting. Our method
would align these features into the following vectors, if $c = 2$: \footnote{In
  fact, the constructor identifier $\id{C}$ would not appear in the vector, but
  this example is still accurate in terms of laying out the features as given.}

@@

To perform this flattening we first transform the nested tokens of an expression
into a \emph{rose tree} of features. Intuitively, these are simply an
s-expression representation of the Core syntax, with the feature extraction
function $\phi$ mapped over the labels. The full transformation is given in
Appendix \ref{sec:rosetree}, and Figure \ref{fig:rosetreeexample} shows the
result for the \hs{odd} function.

@@

@@

These rose trees are then turned into matrices, as shown in Figure
\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features at
depth $i$ in the rose tree, read left-to-right, followed by any required
padding. These matrices are then either truncated, or padded with rows (on the
bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.

\begin{sloppypar}
  Finally, matrices are turned into vectors by concatenating the rows from top
  to bottom, hence Figure \ref{fig:matrixexample} will produce a vector
  beginning
  $(\feature{\CLam} ,
    0               ,
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\id{a}},
    \feature{\CCase},
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\CVar} ,
    \feature{\id{b}},
    \feature{\CAlt} ,
    \feature{\CAlt} ,
    \dots$.
\end{sloppypar}

@@

We now define the function $\phi$, which turns terminal symbols of Core syntax
into features (real numbers). For known language features, such as
$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
assign a value to each, in a similar way to \cite{DBLP:journals/corr/HerasK14}
in Coq. We use a constant $\alpha$ to separate these values from those of other
tokens (e.g. identifiers), but the order is essentially arbitrary: \footnote{In
  \cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina tokens like \coq{fix}
  and \coq{cofix} are grouped together to reduce redundancy; we do not group
  tokens, but we do put ``similar'' tokens close together, such as \CLocal\ and
  \CGlobal.}

@@

To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which
gives equal values for $\alpha$-equivalent expressions (i.e. renaming an
identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
  index} of the identifier \cite{de1972lambda}, denoted $i_l$:

@@

Instead, we use the approach taken in the latest versions of ML4PG and encode
global identifiers \emph{indirectly}, by looking up the expressions which they
\emph{reference}:

@@

For this recursive process to be well-founded, we perform a topological sort on
declarations based on their dependencies (the expressions they reference). In
this way, we can avoid looking up expressions which haven't been clustered
yet. To handle mutual recursion we use Tarjan's algorithm \cite{tarjan1972depth}
to produce a sorted list of \emph{strongly connected components} (SCCs), where
each SCC is a mutually-recursive sub-set of the declarations. If an identifier
cannot be found amongst those clustered so far, it must appear in the same SCC
as the expression we are processing; hence we give that identifier the constant
feature value $f_{recursion}$.

@@

By working through the sorted list of SCCs, storing the features of each
top-level expression as they are calculated, our algorithm can be computed
\emph{iteratively} rather than recursively, as shown in Algorithm
\ref{alg:recurrent}.

As an example of this recurrent process, we can consider the Peano arithmetic
functions from Figure \ref{fig:coreexample}. A valid topological ordering is
given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
Core expressions to save space):

@@

We can then trace the execution of Algorithm \ref{alg:recurrent} as follows:

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$). A more extensive investigation of these features requires a
concrete implementation, in particular to pin down values for the parameters
such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.

@@

We provide an implementation of our recurrent clustering algorithm in a tool
called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
for the GHC compiler, which emits a serialised version of each Core definition
as it is being compiled. This approach is more robust than, for example, parsing
source files, since it avoids the complications of preprocessors and language
extensions altering the syntax.

>>>>

To improve the performance of theory exploration tools, we need to avoid
brute-force search in favour of a more informed strategy. This, in turn,
requires more information about the terms in the given signature. One
generally available source of information is the syntax of a term's definition.

Unfortunately these definitions are over-complicated due to the relatively large
syntax of Haskell, which includes many convenience features (``syntactic
sugar'') and alternative ways to express the same operation (e.g. binding using
\hs{let} or \hs{where}, pattern matching with \hs{if} or \hs{case}, function
abstraction via \hs{f x = ...} or \hs{f = \x -> ...}, etc.). This obscures the
simplicity of the lambda calculus underlying the language, and more importantly
it makes syntactic analysis less useful for our purposes by drawing distinctions
between semantically identical terms. Since theory exploration is concerned only
with semantics (e.g. telling us how code actually behaves), we perform our
analyses on the syntax of the \emph{GHC Core} language instead.

@@

GHC Core is an intermediate language of the \textsc{GHC} compiler based on
\fc{}, and described in detail in~\cite[Appendix C]{sulzmann2007system}. We
ignore type annotations and coercions, since these have no effect on the
semantics (they have no ``computational content''). The resulting sub-set of
Core\footnote{As of GHC version 7.10.2.} is shown in
Figure~\ref{fig:coresyntax}.

@@

GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
shows the Core representation of some simple functions. Although Core is more
verbose, it has fewer redundant features than the full Haskell syntax: for
example all pattern-matching is represented with \texttt{Case}, all function
abstraction uses \texttt{Lam}(bda), and so on. We can see that similar structure
in the Haskell definitions gives rise to similar structure in the Core, such as
the definitions of \hs{odd} and \hs{even} being identical except for the
identifiers. It is this close correspondence which allows us to analyse Core
expressions in place of their more complicated Haskell source.

As a further simplification, we also replace data constructors in definitions
with normal functions, e.g. replacing occurrences of a constructor
\hs{S :: Nat -> Nat} with a function \hs{s} defined as \hs{s x = S x} (this is
known as the $\eta$-expansion of \hs{S}). We do this to remove more distinctions
from the syntax, and also to reduce cross-language differences (in particular,
the Isabelle language used by the \isascheme{}, \isacosy{} and \hipster{} theory
exploration systems treats constructors in a different way to Haskell, so having
constructors in our data would complicate comparisons).

@@

Raw syntax trees, even in Core form, are difficult to analyse due to their
recursive structure, use of discrete labels and heavy reliance on references.
This makes them unsuitable for many machine learning algorithms, which instead
require a fixed-size vector of continuous numbers known as a
\emph{feature vector}.

The process of turning raw data into feature vectors is called \emph{feature
  extraction}, and we adapt the methodology of \emph{recurrent clustering}
proposed in~\cite{DBLP:journals/corr/HerasK14,heras2013proof}, since this
handles recursive tree structures, resolves references and produces numeric
values. We suggest a new recurrent clustering and feature extraction algorithm
for Haskell (shown in Algorithm \ref{alg:recurrent}) and compare its similarity
and differences to the existing approaches of \textsc{ML4PG} and
\textsc{ACL2(ml)}.

@@

We choose k-means because it is a standard approach and we are more concerned
with the application of feature extraction to Haskell and its use in theory
exploration, rather than precise tuning of learning algorithms. There are many
other clustering algorithms we could use, such as \emph{expectation
  maximisation}, but experiments with \textsc{ML4PG} have shown little
difference in their results; it seems that the quality of the features is the
bottleneck to learning, so there is no reason to avoid a fast algorithm like
k-means. We use a Haskell implementation provided by the package
\hPackage{k-means}.

@@

Many conservative improvements can be made to the standard k-means algorithm
described above. For example, a more efficient approach like
\emph{yinyang k-means}~\cite{conf/icml/DingZSMM15} could make larger input sizes
more practical to cluster; the \emph{k-means++}
approach~\cite{arthur2007k, bahmani2012scalable} can be used to more carefully
select the ``seed'' values for the first timestep; and the \emph{x-means}
algorithm~\cite{pelleg2000x} is able to estimate how many clusters to use
(instead of relying on our ``rule of thumb''). However, like the choice of
k-means versus other clustering algorithms, these algorithmic improvements would
need higher quality data than we are capable of producing in these early
experiments.

\subsubsection{Recurrent Clustering}

\emph{Recurrent clustering} has been implemented in the
ML4PG tool for analysing Coq proofs~\cite{journals/corr/abs-1212-3618} and
\textsc{ACL2(ml)} for ACL2~\cite{heras2013proof}. Both approaches transform
syntax trees into fixed-size matrices, then concatenate the rows of these
matrices to produce feature vectors. \textsc{ACL2(ml)} produces matrices by
\emph{summarising} information about the tree; for example, one column counts
the number of variables appearing at each tree level, others count the number of
function symbols which are nullary, unary, binary, etc. In contrast, the
algorithm of ML4PG converts tokens of the syntax tree directly into numbers,
arranging them in the matrix based on their position in the syntax tree. It is
this latter algorithm that we adapt for analysing GHC Core expressions.

The key to recurrent clustering is its treatment of identifiers. From a purely
symbolic perspective, they are atomic: two identifiers can either be identical
or distinct. Consider the \hs{odd} and \hs{even} functions in
Figure~\ref{fig:coreexample}, which differ only in their choice of identifiers.
These functions are clearly different (opposite, in fact), so their identifiers
should play a role in determining their similarity. Yet these distinctions, e.g.
between \hs{True} and \hs{False}, seem intuitively small relative to the
distinction between, for example, \hs{isEmpty} and \hs{isValidEmailAddress}.

Recurrent clustering solves this dichotomy by introducing a finer granularity to
the treatment of identifiers, by first clustering the definitions that they
refer to. If two definitions appear in the same cluster, their identifiers will
be denoted by similar numbers during subsequent feature extraction; if they
appear in different clusters, their numbers will differ more during feature
extraction. It is this use of clustering to determine features that makes the
approach \emph{recurrecnt}.

\subsection{Expressions to Vectors}
\label{sec:expressionstovectors}

The first step of our feature extraction algorithm is to convert the tree
structure of Core expressions to a flat vector. Since k-means clustering
considers the elements of a feature vector to be \emph{orthogonal}, we must
ensure that similar expressions not only give rise to similar numerical values,
but crucially that these values appear \emph{at the same position} in the
feature vectors. Different patterns of nesting can alter the ``shape'' of
expressions, so simple traversals (breadth-first, depth-first, post-order, etc.)
may cause features from equivalent sub-expressions to be mis-aligned. For
example, consider the following Core expressions, which represent pattern-match
clauses with different patterns but the same body (\hs{\vlocal{y}}):

@@

If we traverse these expressions in breadth-first order, converting each token
to a feature using some function $\phi$ (defined later) and padding to the same
length with $0$, we would get the following feature vectors:

@@

The method we use to ``flatten'' expressions, described below, is a variation of
breadth-first traversal which pads each level of nesting to a fixed size $c$
(for \emph{columns}). This doesn't guarantee alignment, but it does prevent
mis-alignment from accumulating across different levels of nesting. Our method
would align these features into the following vectors, if $c = 2$:

@@

@@

To perform this flattening we first transform the nested tokens of an expression
into a \emph{rose tree} of features, via the function $toTree$ defined below.
Intuitively, rose trees are simply an s-expression representation of the Core
syntax, with the feature extraction function $\phi$ mapped over the labels:
Figure~\ref{fig:rosetreeexample} shows the result of transforming the \hs{odd}
function from Figure~\ref{fig:coreexample}.

We follow the presentation in~\cite{blundell2012bayesian} and define rose trees
recursively as follows: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$,
where $f \in \mathbb{R}$ and $T_i$ are rose trees.

The recursive definition of $toTree$ is mostly routine; each repeated element
(shown as $\dots$) has an example to indicate their handling, e.g. for $\CRec$
we apply $toTree$ to each $e_i$. We ignore values of $\mathcal{D}$, since
constructors have no internal structure for us to compare. We also ignore values
from $\mathcal{S}$ and $\mathcal{N}$ as it simplifies our later definition of
$\phi$, and we conjecture that the effect of such ``magic values'' on clustering
real code is low.

\begin{align*}\label{fig:totree}
  toTree(e) &=
  \begin{cases}
    (\feature{\CVar},     toTree(e_1))                                 & \text{if $e = \CVar\ e_1$} \\
    (\feature{\CLit},     toTree(e_1))                                 & \text{if $e = \CLit\ e_1$} \\
    (\feature{\CApp},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CApp\ e_1\ e_2$} \\
    (\feature{\CLam},     toTree(e_1))                                 & \text{if $e = \CLam\ l_1\ e_1$} \\
    (\feature{\CLet},     toTree(e_1), toTree(e_2))                    & \text{if $e = \CLet\ e_1\ e_2$} \\
    (\feature{\CCase},    toTree(e_1), toTree(a_1), \dots)             & \text{if $e = \CCase\ e_1\ l_1\ a_1\ \dots$} \\
    (\feature{\CType})                                                & \text{if $e = \CType$} \\
    (\feature{\CLocal},   (\feature{l_1}))                            & \text{if $e = \CLocal\ l_1$} \\
    (\feature{\CGlobal},  (\feature{g_1}))                            & \text{if $e = \CGlobal\ g_1$} \\
    (\feature{\CConstructor})                                         & \text{if $e = \CConstructor\ d_1$} \\
    (\feature{\CLitNum})                                              & \text{if $e = \CLitNum\ n_1$} \\
    (\feature{\CLitStr})                                              & \text{if $e = \CLitStr\ s_1$} \\
    (\feature{\CAlt},     toTree(e_1), toTree(e_2))                   & \text{if $e = \CAlt\ e_1\ e_2\ l_1\ \dots$}  \\
    (\feature{\CDataAlt})                                             & \text{if $e = \CDataAlt\ g_1$}  \\
    (\feature{\CLitAlt},  toTree(e_1))                                & \text{if $e = \CLitAlt\ e_1$}  \\
    (\feature{\CDefault})                                             & \text{if $e = \CDefault$}  \\
    (\feature{\CNonRec},  toTree(e_1))                                & \text{if $e = \CNonRec\ e_1$}  \\
    (\feature{\CRec},     toTree(e_1), \dots)                         & \text{if $e = \CRec\ e_1\ \dots$} \\
    (\feature{\CBind},    toTree(e_1))                                & \text{if $e = \CBind\ l_1\ e_1$}
  \end{cases}
\end{align*}

@@

These rose trees are then turned into matrices, as shown in
Figure~\ref{fig:matrixexample}. Each row $i$ of the matrix contains the features
at depth $i$ in the rose tree, read left-to-right, followed by any required
padding. These matrices are then either truncated, or padded with rows (on the
bottom) or columns (on the right) of zeros, to fit a fixed size $r \times c$.

\begin{sloppypar}
  Finally, matrices are turned into vectors by concatenating the rows from top
  to bottom, hence Figure~\ref{fig:matrixexample} will produce a vector
  beginning
  $(\feature{\CLam} ,
    0               ,
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\id{a}},
    \feature{\CCase},
    0               ,
    0               ,
    0               ,
    0               ,
    \feature{\CVar} ,
    \feature{\id{b}},
    \feature{\CAlt} ,
    \feature{\CAlt} ,
    \dots$.
\end{sloppypar}

@@

We now define the function $\phi$, which turns terminal symbols of Core syntax
into features (real numbers). For known language features, such as
$\feature{\CLam}$ and $\feature{\CCase}$, we can enumerate the possibilities and
assign a value to each, in a similar way to~\cite{DBLP:journals/corr/HerasK14}
in Coq. We use a constant $\alpha$ to separate these values from those of other
tokens (e.g. identifiers), but the order is essentially
arbitrary:~\footnote{In~\cite{DBLP:journals/corr/HerasK14}, ``similar'' Gallina
  tokens like \coq{fix} and \coq{cofix} are grouped together to reduce
  redundancy; we do not group tokens, but we do put ``similar'' tokens close
  together, such as \CLocal\ and \CGlobal.}

@@

To encode \emph{local} identifiers $\mathcal{L}$ we would like a quantity which
gives equal values for $\alpha$-equivalent expressions (i.e. renaming an
identifier shouldn't affect the feature). To do this, we use the \emph{de Bruijn
  index} of the identifier~\cite{de1972lambda}, denoted $i_l$:

@@

Instead, we use the approach taken nu ML4PG and encode global identifiers
\emph{indirectly}, by looking up the expressions which they \emph{reference}:

@@

For this recursive process to be well-founded, we perform a topological sort on
declarations based on their dependencies (the expressions they reference). In
this way, we can avoid looking up expressions which haven't been clustered
yet. To handle mutual recursion we use Tarjan's algorithm~\cite{tarjan1972depth}
to produce a sorted list of \emph{strongly connected components} (SCCs), where
each SCC is a mutually-recursive sub-set of the declarations. If an identifier
cannot be found amongst those clustered so far, it must appear in the same SCC
as the expression we are processing; hence we give that identifier the constant
feature value $f_{recursion}$.

@@

By working through the sorted list of SCCs, storing the features of each
top-level expression as they are calculated, our algorithm can be computed
\emph{iteratively} rather than recursively, as shown in
Algorithm~\ref{alg:recurrent}.

\begin{figure}
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \begin{tikzpicture}[node distance=2cm]
      \tikzset{vertex/.style = {minimum size=1.5em}}
      \tikzset{edge/.style = {->,> = latex'}}

      \node[vertex]                at (0.5\textwidth, 3)   (z)     {\hs{Z}};
      \node[vertex]                at (0.5\textwidth, 0)   (s)     {\hs{S}};
      \node[vertex]                at (0,             1.5) (plus)  {\hs{plus}};
      \node[vertex, anchor=east]   at (\textwidth,    1.5) (false) {\hs{False}};
      \node[vertex, right of=plus]                         (mult)  {\hs{mult}};
      \node[vertex, left of=false]                         (odd)   {\hs{odd}};
      \node[vertex, left of=odd]                           (even)  {\hs{even}};
      \node[vertex, left of=even]                          (true)  {\hs{True}};

      \draw[edge, dashed]     (z)     to (plus);
      \draw[edge, dashed]     (z)     to (mult);
      \draw[edge, dashed]     (z)     to (odd);
      \draw[edge, dashed]     (z)     to (even);
      \draw[edge, dashed]     (s)     to (plus);
      \draw[edge, dashed]     (s)     to (mult);
      \draw[edge, dashed]     (s)     to (odd);
      \draw[edge, dashed]     (s)     to (even);
      \draw[edge, dashed]     (false) to (odd);
      \draw[edge, dashed]     (true)  to (even);
      \draw[edge]             (plus)  to (mult);
      \draw[edge]             (even)  to (odd);
      \draw[edge]             (odd)   to (even);
      \draw[edge, loop above] (plus)  to (plus);
      \draw[edge, loop above] (mult)  to (mult);
    \end{tikzpicture}

    \caption{Dependency graph for Figure \ref{fig:coreexample}. Loops indicate
      recursive functions, double arrows indicate mutual recursion. Dashed lines
      show references to data constructors, which we do not consider in this
      work.}
    \label{fig:dagexample}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \centering
    \tikzstyle{block} = [rectangle, draw]
    \tikzstyle{line}  = [draw, -latex']
    \begin{tikzpicture}[node distance=2cm]
      \node [block] (plusset) {
        \begin{tikzpicture}[node distance = 2cm]
          \node [block] (plus) {\hs{plus}};
        \end{tikzpicture}
      };
      \node [block, right of=plusset] (parityset) {
        \begin{tikzpicture}[node distance = 1cm]
          \node [block]               (odd)  {\hs{odd}};
          \node [block, below of=odd] (even) {\hs{even}};
        \end{tikzpicture}
      };
      \node [block, right of=parityset] (multset) {
        \begin{tikzpicture}[node distance = 2cm]
          \node [block] (mult) {\hs{mult}};
        \end{tikzpicture}
      };

      \path [line] (plusset)   -- (parityset);
      \path [line] (parityset) -- (multset);
    \end{tikzpicture}
    \caption{One possible topological sorting of simply connected components for
      Figure \ref{fig:dagexample} (ignoring constructors). \hs{even} and
      \hs{odd} are mutually recursive, neither can appear before the other, so
      they are grouped into one component and handled concurrently by Algorithm
      \ref{alg:recurrent}.}
    \label{fig:sccexample}
  \end{subfigure}
  \caption{Sorting functions from Figure \ref{fig:coreexample} into dependency order.}
  \label{fig:dependencyexample}
\end{figure}

As an example of this recurrent process, we can consider the Peano arithmetic
functions from Figure~\ref{fig:coreexample}. A valid topological ordering is
given in Figure~\ref{fig:sccexample}, which can be our value for $d$ (eliding
Core expressions to save space):

@@

We can then trace the execution of Algorithm~\ref{alg:recurrent} as follows:

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$).

@@

We provide an implementation of our recurrent clustering algorithm in a tool
called \mlforhs{}. We obtain Core ASTs from Haskell packages using a plugin
for the GHC compiler~\footnote{Available at
  \url{https://github.com/warbo/ast-plugin}}, which emits a serialised version
of each Core definition as it is being compiled. This approach is more robust
than, for example, parsing source files, since it avoids the complications of
preprocessors and language extensions altering the syntax.
